Taking Dyads Seriously

### 2018-01-04 paper/2_model.tex
added discussion of symmetry, changed stuff about R&S/Rose

The presence of these dependencies in relational data points to the fact that there is a complex structure underlying the dyadic events that we observe, and that accounting for this structure is necessary if we are to adequately represent the data generating process. Of course, if one can specify each of the nodal, dyadic, and triadic set of attributes that influence interactions then one can be assured that the conditional independence assumption underlying standard approaches will be satisfied. However, it is rarely the case that we can do this even for TSCS data, thus we more than often make modeling decisions to account for that structure. Failing to account for the underlying structure in either TSCS or dyadic data leads to a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. Further by ignoring these potential interdependencies, we often ignore substantively interesting features of the problem. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?

@@

\begin{align}
	\begin{aligned}
		y_{ij} \;=\; f(\theta_{ij}) &\text{, where } \\
		\theta_{ij} \;=\;& \bm\beta_{d}^{\top} \mathbf{X}_{ij} + \bm\beta_{s}^{\top} \mathbf{X}_{i} + \bm\beta_{r}^{\top} \mathbf{X}_{j} \text{\qquad(Exogenous parameters)} \\
		& + a_{i} + b_{j} + \epsilon_{ij} \text{\qquad\qquad\qquad\quad(SRRM parameters)} \\
		& + \mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}  \text{\qquad\qquad\qquad\qquad\qquad\;(LFM parameters)} \\ 
	\label{eqn:ame}
	\end{aligned}
\end{align}

where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver). We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, using first a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$). Next, to account for the dependencies that emerge in dyadic data and that may complicate inference on the parameter associated with exogenous covariates, we add parameters from the SRRM and LFM. $a_{i}$ and $b_{j}$ in Equation~\ref{eqn:ame} represent sender and receiver random effects incorporated from the SRRM framework:

@@

The LFM contribution to the AME comes in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies \citep{minhas:etal:2016:arxiv}. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.}

>>>>

The presence of these dependencies in relational data points to the fact that there is a complex structure underlying the dyadic events that we observe, and that accounting for this structure is necessary if we are to adequately represent the data generating process. Of course, if one can specify each of the nodal, dyadic, and triadic set of attributes that influence interactions then one can be assured that the conditional independence assumption underlying standard approaches will be satisfied. However, it is rarely the case that we can do this even for TSCS data, thus we often make modeling decisions to account for that structure. Failing to account for the underlying structure in either TSCS or dyadic data leads to a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. Further by ignoring these potential interdependencies, we often ignore substantively interesting features of the problem. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?

@@

\begin{align}
	\begin{aligned}
		y_{ij} \;=\; f(\theta_{ij}) &\text{, where } \\
		\theta_{ij} \;=\;& \bm\beta_{d}^{\top} \mathbf{X}_{ij} + \bm\beta_{s}^{\top} \mathbf{X}_{i} + \bm\beta_{r}^{\top} \mathbf{X}_{j} \text{\qquad(Exogenous parameters)} \\
		& + a_{i} + b_{j} + \epsilon_{ij} \text{\qquad\qquad\qquad\quad(SRRM parameters)} \\
		& + \mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}  \text{\qquad\qquad\qquad\qquad\qquad\;(LFM parameters)} \\ 
	\label{eqn:ame}
	\end{aligned}
\end{align}
where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver). We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, using first a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$). Next, to account for the dependencies that emerge in dyadic data and that may complicate inference on the parameter associated with exogenous covariates, we add parameters from the SRRM and LFM. $a_{i}$ and $b_{j}$ in Equation~\ref{eqn:ame} represent sender and receiver random effects incorporated from the SRRM framework:

@@

The LFM contribution to the AME comes in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies \citep{minhas:etal:2016:arxiv}. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.} Note that this model easily generalizes to the case, common in international relations, where interactions are unidrected (for example the presence of conflict or a bilateral investment treaty). In this case, $\rho$ is constrained to be 1, sender and receiver effects are identical for each observation, as are \mathbf{u_{i}}$ and $\mathbf{v_{j}}$. In the application section, we show the applicability of the AME approach to both directed and undirected dyadic data.

### 2018-01-04 paper/4_app.tex
stuff
added discussion of symmetry, changed stuff about R&S/Rose

\begin{table}[ht]
\centering
\caption{Here we provide a brief summary of the key variable in each of the five replications and a note about whether or not the finding is replicated when using our network-based approach.}
	\begin{tabular}{l p{7cm} l} \toprule
		\multirow{2}{*}{Study} & \multirow{2}{*}{Central Finding} &  Replicates \\ 
			& &  in a Network Model? \\ \toprule
		Reiter \& Stam (2003) & Personalist Regimes Attack Democracies, Not Vice Versa & {Does Not Replicate} \\ \midrule
		McDonald (2004) & Lower Trade Barriers and Higher Trade Lead to Peace & {Does Not Replicate}\\ \midrule
		Rose (2004) & WTO Membership Does not Affect Trade & {Partially Replicates}\\ \midrule
		Weeks (2012) & Bosses, Juntas, and Strongmen are more Aggressive, Machines are Not & {Does Not Replicate} \\\midrule
		Gibler (2017) & Power Parity at Time of Entry to International System Inceases Conflict & {Does Not Replicate}\\ \bottomrule
	\end{tabular}
	\label{tab:modelFindingSumm}
\end{table}

@@

In the original model, the variable ``Pers/Democ Directed Dyad" (which represents a Personalist $\rightarrow$ Democractic directed dyad) is clearly positive while the variable ``Democ/Personalist Directed Dyad'' is zero and the difference between the two coefficients is clearly distinct from zero. In our re-estimation using the AME framework, we also find that Pers-Democ directed dyad has a positive effect with zero excluded from the 95\% confidence interval while Democ-Pers directed dyad is indistinguishable from zero. Using this model, however, we can no longer conclusively say that the Pers/Democratic coefficient is larger than the Democ/Personalist one. Our re-estimation using the AME approach therefore cast some doubt on Reiter \& Stam's key claim that MIDs initiated by personalist dictatorships against democracies are more likely than MIDS initiated by democracies. Further, the effect of most of the covariates in the literature thought to predict interstate MIDs are much closer to zero when using the AME framework. An important takeaway here is that many studies make knowledge claims based on the statistical significance of a small set of covariates, or the differences between these covariates. These differences may change dramatically when interdependencies are taken into account directly, as they do in the case of this study.

@@

When we compare the results of Rose's original OLS model to the AME model accounting for network dependencies, the results are similar. The main result of the model---the null effect of membership in the WTO, as represented by the ``One-In" and ``Both-In" variables---does not prevail in the AME model. In the original, this negative linkage is not significant, but in the AME estimation this negative linkage is more precisely estimated. However, while in the original model there was a clear positive relationship between Real GDP and Trade, most of this effect vanishes in the AME model. The random effects shown in Figure~\ref{fig:roser} reveal the cause of much of this divergence. Here, the states with the most positive random effects are also states with high GDP, though not necessarily high GDP/capita.\footnote{Note: Qatar exhibits strongly negative random effects.} Thus, the effect of GDP in the original model was, in part, an artifact of first-order dependencies. Most of the other results of the model are constant across each model, though some geographic features, such as islands and landlocked states, have a more clear effect on trade once we account for these network dependencies. When we account for network interdependencies, we observe a markedly lower Root Mean Squared Error out-of-sample --- $3.23$ for the OLS model and $1.77$ for the AME model. While the AME model replicate the main result, there are substantial and substantive improvements gleaned from the AME results.

@@

\begin{figure}[!h]
	\includegraphics[width=\textwidth]{weeks_circPlot.pdf}
	\caption{\label{fig:weekscirc} Visualization of multiplicative effects for Weeks (2012). Each circle designates a country and the color corresponds to the legend at the center of the visualization. Countries that cluster together in the outer ring are those that were found by the model to have similar sending patterns, meaning that they tend to send conflict to similar sets of countries. The inner ring clusters countries by the similarity of who they receive conflict from. 
	%Blue represents groups with common sending patterns and red represents groups with common receiving patterns.
	}
\end{figure}
\FloatBarrier

>>>>

\begin{table}[ht]
\centering
\caption{Here we provide a brief summary of the key variable in each of the five replications and a note about whether or not the finding is replicated when using our network-based approach.}
	\begin{tabular}{l p{7cm} l} \toprule
		\multirow{2}{*}{Study} & \multirow{2}{*}{Central Finding} &  Replicates \\ 
			& &  in a Network Model? \\ \toprule
		Reiter \& Stam (2003) & Personalist Regimes Attack Democracies, Not Vice Versa & {Partially Replicates} \\ \midrule
		McDonald (2004) & Lower Trade Barriers and Higher Trade Lead to Peace & {Does Not Replicate}\\ \midrule
		Rose (2004) & WTO Membership Does not Affect Trade & {Partially Replicates}\\ \midrule
		Weeks (2012) & Bosses, Juntas, and Strongmen are more Aggressive, Machines are Not & {Does Not Replicate} \\\midrule
		Gibler (2017) & Power Parity at Time of Entry to International System Inceases Conflict & {Does Not Replicate}\\ \bottomrule
	\end{tabular}
	\label{tab:modelFindingSumm}
\end{table}

@@

In the original model, the variable ``Pers/Democ Directed Dyad" (which represents a Personalist $\rightarrow$ Democractic directed dyad) is clearly positive while the variable ``Democ/Personalist Directed Dyad'' is zero and the difference between the two coefficients is clearly distinct from zero. In our re-estimation using the AME framework, we also find that Pers-Democ directed dyad has a positive effect with zero excluded from the 95\% confidence interval while Democ-Pers directed dyad is indistinguishable from zero. Using this model, however, we can no longer conclusively say that the Pers/Democratic coefficient is larger than the Democ/Personalist one. Our re-estimation using the AME approach therefore cast some doubt on Reiter \& Stam's key claim that MIDs initiated by personalist dictatorships against democracies are more likely than MIDS initiated by democracies, though many of the effects uncovered by the authors remains robust when accounting for dyadic interdependencies. Further, the effect of most of the covariates in the literature thought to predict interstate MIDs are much closer to zero when using the AME framework. %An important takeaway here is that many studies make knowledge claims based on the statistical significance of a small set of covariates, or the differences between these covariates. These differences may change dramatically when interdependencies are taken into account directly, as they do in the case of this study.

@@

When we compare the results of Rose's original OLS model to the AME model accounting for network dependencies, the results are similar. The main result of the model---the null effect of membership in the WTO, as represented by the ``One-In" and ``Both-In" variables---does not prevail in the AME model. Howevever, in a sense the AME result provides even stronger evidence against a positive effect of WTO membership: in the original, this negative linkage is not significant, but in the AME estimation this negative linkage is more precisely estimated. In both cases we find  a rejection of the conventional wisdom, that WTO membership increases trade. While in the original model there was a clear positive relationship between Real GDP and Trade, most of this effect vanishes in the AME model. The random effects shown in Figure~\ref{fig:roser} reveal the cause of much of this divergence. Here, the states with the most positive random effects are also states with high GDP, though not necessarily high GDP/capita.\footnote{Note: Qatar exhibits strongly negative random effects.} Thus, the effect of GDP in the original model was, in part, an artifact of first-order dependencies. Most of the other results of the model are constant across each model, though some geographic features, such as islands and landlocked states, have a more clear effect on trade once we account for these network dependencies. When we account for network interdependencies, we observe a markedly lower Root Mean Squared Error out-of-sample --- $3.23$ for the OLS model and $1.77$ for the AME model. While the AME model replicates the main (null) result, there are substantial and substantive improvements gained from moving to an AME estimation.

@@

\begin{figure}[!h]
	\includegraphics[width=\textwidth]{weeks_circPlot.png}
	\caption{\label{fig:weekscirc} Visualization of multiplicative effects for Weeks (2012). Each circle designates a country and the color corresponds to the legend at the center of the visualization. Countries that cluster together in the outer ring are those that were found by the model to have similar sending patterns, meaning that they tend to send conflict to similar sets of countries. The inner ring clusters countries by the similarity of who they receive conflict from. 
	%Blue represents groups with common sending patterns and red represents groups with common receiving patterns.
	}
\end{figure}
\FloatBarrier
